Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-25 appendix.tex
update bias plot/discussion
going thru comments
update notation

Assume $w=1$.
In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the divergence term is identical for $\tau_1$ and $\tau_2$ as when the internal branch becomes zero-length the two topologies are indistinguishable.
Let $\mathcal{T}$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
The unconstrained maximum likelihood value for $\tau_2$ is greater than or equal to the value of the likelihood given these two constraints.
We have
\begin{align*}
\max_{t} & \ \ell_{\tau^*,t^*}(\tau_2, t) \ge \\
&\max_{t: \xi_j=\emptyset, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}) \\
&= \max_{t: \xi_j=\emptyset, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t : t^*\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau_1, t),
\end{align*}
and thus the incorrect topology achieves a likelihood value greater than or equal to the correct one.

@@

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{empirical-bias}
\caption{
Bias in branch length estimation.
In regions where joint optimization should perform well, there is systematic bias toward shorter branch lengths.
}
\label{fig:empirical-bias}
\end{figure}

>>>>

Assume $w=1$.
In this case, Table~\ref{tab:gen-sitepatprob} shows that, for $w=1$, the divergence term is identical for $\tau_1$ and $\tau_2$ as when the internal branch becomes zero-length the two topologies are indistinguishable.
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the shaded region in Fig.~\ref{fig:incons-analytic}.
The unconstrained maximum likelihood value for $\tau_2$ is greater than or equal to the value of the likelihood given these two constraints.
We have
\begin{align*}
\max_{t} & \ \ell_{\tau^*,t^*}(\tau_2, t) \ge \\
&\max_{t: \xi_j=\emptyset, w=1, \tau=\tau_2} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}) \\
&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}) \\
&= \max_{t: \xi_j=\emptyset, w=1, \tau=\tau_1} \ \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&\qquad\qquad + \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t : t^*\in\mathcal{T}_0} \ \ell_{\tau^*,t^*}(\tau_1, t),
\end{align*}
and thus the incorrect topology achieves a likelihood value greater than or equal to the correct one.

@@
